Inference using context-dependent models of mutation

### 2017-04-27 writeups/context-dependence.tex
switched fit-model to full likelihood function
some more words

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage[colorlinks=false]{hyperref}
\usepackage{bm}
\usepackage{todonotes}

@@

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\calS}{\mathcal{S}}  % set of states
\newcommand{\calT}{\mathcal{T}}  % set of transition triples
\newcommand{\nA}{\mbox{A}}  % nucleotides:
\newcommand{\nC}{\mbox{C}}
\newcommand{\nG}{\mbox{G}}
\newcommand{\nT}{\mbox{T}}
\newcommand{\join}{\oplus}  % matches
\newcommand{\st}{\colon}  % such that
\newcommand{\var}{\mathop{\mbox{var}}}
\newcommand{\cov}{\mathop{\mbox{cov}}}
\newcommand{\given}{\;\mid\;}
\newcommand{\like}{\mathcal L}
\newcommand{\loglike}{\ell}
\newcommand{\alike}{\widetilde{\like}}
\newcommand{\aloglike}{\widetilde{\loglike}}

@@

To formalize this notion of local dependence,
first define a \textit{pattern} to mean a contiguous sequence of states;
let $|u|$ denote the length of the pattern $u$.
Given $x \in \calS$,
let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denote the pattern of $x$ of length $h$ beginning at location $i$.
Recall we are focusing on finite sequences, and by default, we do not allow patterns to ``hang off'' the end of the sequence.
This is natural in many situations, but in others we might want to other boundary conditions,
modeling extant, but unobserved, neighbors.
We will mention this when necessary.

The dynamics of our stochastic process are determined by the set of allowed transitions and associated rates:
given two patterns $u$ and $v$ of common length $h$,
saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that
\[
    \P\{ X_i^{(h)}(t+dt) = v \given X_i^{(h)}(t) = u \} = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
Such a rule is defined by its ``transition triple'' $(\mu,u,v)$,
written as ``$u \to v$ at rate $\mu$''.
The set of all transition triples, $\calT$, determines the time evolution of this process.

\peter{INCLUDE SELECTION}

@@

\begin{example}[CpG mutation]
  A sequence of genome can be written using A, C, G, and T;
  in the most general model of independent mutation across sites, each of the 12 possible transitions occurs at its own rate.
  Furthermore, it is a well-known observation that in many species,
  adjacent, methylated CG dinucleotides (``CpG sites'') have a much higher mutation rate to TG and CA
  than either single nucleotide change under the independent model.
  Embellishing the single-nucleotide model with this additional rate results in the model defined by

@@

\begin{example}[1-D Ising model with Glauber dynamics]
  In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
  imagined as magnetic dipoles,
  and the energy associated with a given state $x$ is $H(x) = - \frac{1}{2} \beta \sum_i x_i x_{i+1} - \frac{1}{2} \gamma \sum_i x_i$.
  Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
  The associated stationary distribution on configurations is proportional to $\exp(-H(x))$.
  The following ``Glauber dynamics'' preserve the stationary distribution:
  each site, independently at rate $\lambda$,
  forgets its spin,
  and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration.
  Ignoring transitions that don't change the state,
  if patterns $u$ and $v$ differ at only one site then $u$ changes to $v$ at rate $\lambda/(1+\exp(H(v)-H(u)))$,
  so

  Glauber dynamics have been widely studied... XXX

@@

As noted above, there may be more than one way to mutate a sequence $x$ to get $y$:
for each position $i$, let $J(i,x,y)$ be those transitions in $\calT$ that can be applied at $i$
to change $x$ into $y$, i.e.
\[
J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}.
\]
The rate $G(n)_{x,y}$ is then the sum of all matching transition rates,
namely
\begin{align} \label{eqn:G_defn}
  G(n)_{x,y} = \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
This formalizes the statement that the transition rate from $x$ to $y$ is the sum of the mutation rates of all transition triples
that would change $x$ into $y$.
In principle, this gives us the transition probabilities for the process by a matrix exponential:
\begin{align} \label{eqn:full_likelihood}
    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = \left(e^{tG(n)}\right)_{xy} ,
\end{align}
which would then provide a route to parameter estimation.
In practice, the size of these matrices makes direct application obviously impractical.
This paper proposes an alternate strategy.

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm''.
Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu_*$ per site,
and then resolving each possible change in temporal order.
The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu_j \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
\]
Note that since mutation is homogeneous, the maximum rate of changes occurring at any position will be the same, except possibly near the end of the sequence,
so the maximization over $i$ is unnecessary.
Now sample possible changes: let $c_{ij}$ denote the time of the $j^\text{th}$ possible change at site $i$,
so that for each site $i$,
\[
0 \le c_{i1} < c_{i2} < \cdots < c_{iN_i} \le t
\]
is a draw from a rate $\mu_*$ Poisson process on $[0,t]$.
These processes are independent across sites.
Now, suppose we have determined the state at time $s$ to be $X(s) = x$,
and the next event is $c_{ij} = \min_{k,h} \{ c_{kh} : c_{kh}>s \}$.
At $c_{ij}$,
a new state is then chosen at position $i$
with probability proportional to the corresponding rate, i.e.\
\erick{Previously, the $j$ in $\mu^j$ was used to index the possible patterns. Now it's indexing possible changes at a site.}
\[
q_j = \begin{cases}
  \mu_j/\mu_* \qquad & \text{if } x_i^{(|u^j|)} = u^j  \\
  0 \qquad & \text{otherwise,}
\end{cases}
\]
is the probability that $x_i^{(|u^j|)}$ is replaced with $v^j$.
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$ gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0\ge 0$.)

@@

The problem at hand is to infer the parameters of the model, given the state at time 0 and later at time $t$.
If $n$ is large enough and $t$ is not too large,
this should be feasible -- there should be enough information in our observations to do this reliably.
If $t$ is too large, the process may ``saturate'' --
if there are many changes at most sites, then in at least most models,
we will lose most information about the dynamics,
retaining information only about the parameters that affect the stationary distribution.
In TASEP with any boundary conditions, we lose all information as it approaches stationarity,
while in the Ising model we have information about the temperature $\beta$ and magnetic field $\gamma$ but not the speed of changes, $\lambda$.
In the CpG model it is not immediately clear what information is retained.

But, how can we extract the information?
These are Markov processes, on the state space $\calS^n$,
so the full likelihood function is given by \eqref{eqn:full_likelihood},
but doing anything with the $|\calS|^n \times |\calS|^n$ matrix $G_n$ is clearly infeasible for even moderately sized $n$.
We can, however, compute \eqref{eqn:full_likelihood} for smaller $n$,
so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
and treat these as independent.
Then, defining $p_m(t;x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
we would obtain the approximate likelihood function
\[
  \prod_{k=0}^{n/m} p_m(x_{km+1}^{(m)},y_{km+1}^{(m)}) .
\]
However, there is no guarantee that this is even close to correct,
as it ignores dependencies between neighboring blocks.
Intuitively, we want to increase the size of the context
-- and the likelihood thus obtained is asymptotically correct for $m$ large,
but computationally, we are stuck with small values of $m$.

Alternatively, we could take into account context in the initial sequence only,
by including the states of an additional $\ell$ sites on both the left and right.
The probability that an initial sequence $x$ of length $m+2\ell$ after time $t$
is found to match a smaller, middle sequence $y$ of length $m$ is found by marginalizing:
\begin{align}
  p_{\ell,m}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^\ell} p_{2\ell+m}(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^{2\ell+m}, \; y \in \calS^m ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
This is the probability that if we begin with a sequence $x$ of length $2\ell+m$,
then after time $t$ we will see the pattern $y$ at the middle $m$ positions,
ignoring what is present in the leftmost and rightmost $\ell$ positions.
We expect $p_{\ell,m}(x,y)$ to be asymptotically correct when $\ell$ is large (and $t$ is fixed),
i.e.\ that the probability a given subsequence matching $x$ is later observed to have its center matching $y$
is close to $p_{\ell,m}(x,y)$, regardless of context wider than $2k+l$:
\begin{align} \label{eqn:window_approx}
  p_{\ell,m}(x,y) \approx \P\{ X_{i+\ell}^{(m)}(t) = y \mid X_i^{(2\ell+m)}(0) = x \}.
\end{align}
We provide a precise statement of the error in this approximation later.

\peter{how to choose window size}

@@

\peter{shorten/edit this to follow on from the above}

@@

\section{Discussion}

@@

%%%%
\section*{Proof of the approximation}

@@

In fact, the approximation improves exponentially as $\ell$ and $m$ increase --
here is a simple argument.
Suppose the maximum length of any transition pattern is $R = \max\{ |u| : (\mu,u,v) \in \calT \}$.
The transition rate at site $i$ at time $s$ is determined if we know $X_{i-R}^{(2R+1)}(s)$;
this can be determined from $X_{i-R}^{2R+1}(0)$ and the outcome of all changes happening before time $s$ at sites in $(i-R-1, \ldots, i+R)$.
These will depend in turn on the outcome of previous possible changes.
For each site $i$ and times $s_1 < s_2$, define the set of states $H(i,s_1,s_2)$
so that if one knows the possible change points $s_{ij}$,
then changes outside of $H(i,s_1,s_2)$ do not affect the distribution of $X_i(s_2)$.
Concretely, define $H(i,s_1,s_2)$ by the property that $i \in H(i,s_1,s_2)$, and
\begin{gather}
  \text{if } s_1 < s_{jk} \le s_2 \text{ for some } i-R-1 \le j \le i+R, \text{ then } H(j,s_1,s_{jk}) \subseteq H(i,s_1,s_2) .
\end{gather}
In other words, $H(i,s_1,s_2)$ is the union of the $R$-neighborhoods of $i$ and all possible changes occurring in $H(i,s_1,s_2)$ between $s_1$ and $s_2$.
For a set of states $I$, let $H(I,s_1,s_2) = \bigcup_{i\in I} H(i,s_1,s_2)$.
Formally, the state $X_i(s_2)$ is conditionally independent of $\{X_j(s_1) \st j \notin H(i,s_1,s_2)\}$.
Informally, the approximation \eqref{eqn:window_approx} will be good if $H$ does note expand beyond the flanking context.
Now, MAKE THIS EASIER TO READ
\begin{multline}
  \P\{ X_i^{(m)}(t) = y \mid X(0) \}
  = \P\left\{ H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
   \qquad \qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
     \qquad {} + \P\left\{ H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\}  \\
   \qquad\qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\} ,
\end{multline}
and so
\begin{align} \label{eqn:prob_approx}
  \left| \P\{ X_i^{(m)}(t) = y \mid X(0) \} - p_{\ell,m}(t;x,y) \right| \le 2 \P\left\{  H(\{I,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\}.
\end{align}

@@

@@

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage[hidelinks]{hyperref}
\usepackage{bm}
\usepackage{todonotes}

@@

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\calS}{\mathcal{S}}  % set of states
\newcommand{\calT}{\mathcal{T}}  % set of transition triples
\newcommand{\calP}{\mathcal{P}}  % set of selection pairs
\newcommand{\nA}{\mbox{A}}  % nucleotides:
\newcommand{\nC}{\mbox{C}}
\newcommand{\nG}{\mbox{G}}
\newcommand{\nT}{\mbox{T}}
\newcommand{\join}{\oplus}  % matches
\newcommand{\st}{\colon}  % such that
\newcommand{\var}{\mathop{\mbox{var}}}
\newcommand{\cov}{\mathop{\mbox{cov}}}
\newcommand{\given}{\;\mid\;}
\newcommand{\like}{\mathcal L}
\newcommand{\loglike}{\ell}
\newcommand{\alike}{\widetilde{\like}}
\newcommand{\aloglike}{\widetilde{\loglike}}

@@

To formalize this notion of local dependence,
first define a \textit{pattern} to mean a contiguous sequence of states;
let $|u|$ denote the length of the pattern $u$.
Given $x \in \calS$,
let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denote the pattern of $x$ of length $h$ beginning at location $i$.
% Recall we are focusing on finite sequences, and by default, we do not allow patterns to ``hang off'' the end of the sequence.
% This is natural in many situations, but in others we might want to other boundary conditions,
% modeling extant, but unobserved, neighbors.
% We will mention this when necessary.
The dynamics of our stochastic process are determined by the set of allowed transitions and associated rates:
given two patterns $u$ and $v$ of common length $h$,
saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that
\[
    \P\{ X_i^{(h)}(t+dt) = v \given X_i^{(h)}(t) = u \} = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
Such a rule is defined by its ``transition triple'' $(\mu,u,v)$,
written as ``$u \to v$ at rate $\mu$''.
The set of all transition triples is denoted $\calT$.

@@

\peter{Should call this ``selection'' instead of a ``potential''? Or what?}

To allow more compact specification of models,
we also introduce a ``potential'': 
suppose that there is a collection of patterns $\{p_i\}$
such that each occurrence of $p_i$ adds a quantity $e_i$ to the total ``energy'' of a sequence,
and that the rate at which each possible change occurs is modulated by a function of the energy difference the change would produce.
This is natural if that transition triples describe \emph{proposed} changes,
and that the probability a proposed change actually occurs depends on how much it affects the energy.
Concretely, let $\calP = \{(p_i, e_i\}$ be a set of (pattern, energy change) pairs,
and for any sequence $x$ define
$E(x) = \sum_i e_i n(x,p_i)$,
where $n(x,p_i)$ is the number of times the pattern $p_i$ occurs in $x$.

\begin{example}[Genomic GC content]
    A sequence of genome can be written using A, C, G, and T;
    in the most general model of independent mutation across sites, each of the 12 possible transitions occurs at its own rate.
    Furthermore, it is a well-known observation that in many species,
    adjacent, methylated CG dinucleotides (``CpG sites'') have a much higher mutation rate to TG and CA
    than either single nucleotide change under the independent model.
    Embellishing the single-nucleotide model with this additional rate results in the model defined by
    \begin{center}
      \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
        $x$  &  $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
        \nC\nG   &  \nT\nG   &  $\gamma$ & \\
        \nC\nG   &  \nC\nA   &  $\gamma$ &
      \end{tabular} ,
    \end{center}
    where $\gamma$ is the additional CpG rate above the base mutation rate.
    Note that for a given pair of sequences, there may be more than one way to mutate a given string $x$ to get $y$:
    for instance, a change $\nA\nC\nG \to \nA\nT\nG$ could have occurred by a single $\nC \to \nT$ mutation,
    or by a $\nC\nG \to \nT\nG$ mutation;
    and hence the total instantaneous rate at which ACG changes to ATG is $\gamma + m_{\nC\nT}$.
    % Although this means there are many possible parameterizations,
    % note that the model is identifiable, not overparameterized.

    Counteracting this trend towards increased G/C is GC-biased gene conversion \citep{gcbiased},
    which is effectively a selective pressure against G or C bases.
    We model the evolution of a \emph{single} sequence, not a population of sequences,
    imagining this sequence to be the consensus sequence of the population.
    A new mutation with an effect $s$ on fitness that occurs
    in a population of effective size $N$ becomes ubiquitous, rather than dying out,
    with probability approximately $(1-\exp(-2 s))/(1-\exp(-2 s N))$ \citep{fixation_prob}.
    If the mutation occurs at rate $\mu$ per individual, the total rate it occurs at in the population is $\mu N$.
    GC-biased gene conversion effectively means that sequences with more G/Cs are selected against.
    This is incorporated as follows: set
    \begin{center}
        \begin{tabular}{cc}
        pattern  &  energy \\
        \nC{} or \nG   &  $s$   
      \end{tabular} ,
    \end{center}
    so that $E(y)-E(x)$ is the net change in number of G's and C's multiplied by $s$.
    Also, let
    \begin{align*}
        \phi(e) = N (1-\exp(-2e))/(1-\exp(-2eN))
    \end{align*}
    so that if a given change that occurs with rate $\mu$ would change a sequence $x$ to $y$,
    then $\mu \phi(E(y)-E(x))$ is the rate at which the mutation appears and successfully takes over in the population.

@@

\begin{example}[1-D Ising model with Glauber dynamics]
    In the Ising model, each site is labeled as either ``up'' or ``down'' ($+1$ or $-1$ respectively),
    imagined as magnetic dipoles,
    and the energy associated with a given state $x$ is $E(x) = - \frac{1}{2} \beta \sum_i x_i x_{i+1} - \frac{1}{2} \gamma \sum_i x_i$.
    Here $\beta$ represents inverse temperature, and $\gamma$ represents the strength of the magnetic field (here scaled by temperature).
    The associated stationary distribution on configurations is proportional to $\exp(-E(x))$.
    The following dynamics preserve the stationary distribution:
    each site, independently at rate $\lambda$,
    forgets its spin,
    and reconfigures to a state chosen with probability proportional to the stationary probability of the resulting configuration:
    Since $\exp(-E(x))/(\exp(-E(x))+\exp(-E(y))) = 1/(1+exp(-(E(x)-E(y)))$,
    this is:
    \begin{center}
        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
          $+$  &   $-$   &  $\lambda$ \\
          $-$  &   $+$   &  $\lambda$ 
        \end{tabular} 
        \qquad and \qquad
        \begin{tabular}{cc}
        pattern  &  energy \\
        $+-$ or $-+$  &   $\beta$ \\
        $+$ &   $\gamma$
        \end{tabular} 
    \end{center}
    and 
    \begin{align*}
        \phi(e) = \frac{1}{1+\exp(-e)} .
    \end{align*}

    This process has been studied by XXX.
\end{example}

@@

As noted above, there may be more than one way to mutate a sequence $x$ to get $y$:
for each position $i$, let $J(i,x,y)$ be those transitions in $\calT$ that can be applied at $i$
to change $x$ into $y$, i.e.
\[
J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}.
\]
The rate $G(n)_{x,y}$ is then the sum of all matching transition rates,
namely
\begin{align} \label{eqn:G_defn}
    G(n)_{x,y} = \sum_{i=1}^n \phi(E(y)-E(x)) \sum_{J(i,x,y)}  \mu^j ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
This formalizes the statement that the transition rate from $x$ to $y$ is the sum of the rates of all transition triples
that would change $x$ into $y$.
In principle, this gives us the transition probabilities for the process by a matrix exponential:
\begin{align} \label{eqn:full_likelihood}
    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = \left(e^{tG(n)}\right)_{xy} ,
\end{align}
which would then provide a route to parameter estimation.
In practice, the size of these matrices makes direct application obviously impractical.
This paper proposes an alternate strategy.

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm''.
Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu_*$ per site,
and then resolving each possible change in temporal order.
The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu_j \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
\]
Note that since the process is homogeneous, the maximum rate of changes occurring at any position will be the same, except possibly near the end of the sequence,
so the maximization over $i$ is unnecessary.
Now sample possible changes: let $c_{ij}$ denote the time of the $j^\text{th}$ possible change at site $i$,
so that for each site $i$,
\[
0 \le c_{i1} < c_{i2} < \cdots < c_{iN_i} \le t
\]
is a draw from a rate $\mu_*$ Poisson process on $[0,t]$.
These processes are independent across sites.
Now, suppose we have determined the state at time $s$ to be $X(s) = x$,
and the next event is $c_{ij} = \min_{k,h} \{ c_{kh} : c_{kh}>s \}$.
At $c_{ij}$,
a new state is then chosen at position $i$
with probability proportional to the corresponding rate, i.e.\
\erick{Previously, the $j$ in $\mu^j$ was used to index the possible patterns. Now it's indexing possible changes at a site.}
\[
q_j = \begin{cases}
  \mu_j/\mu_* \qquad & \text{if } x_i^{(|u^j|)} = u^j  \\
  0 \qquad & \text{otherwise,}
\end{cases}
\]
is the probability that $x_i^{(|u^j|)}$ is replaced with $v^j$.
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$ gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0\ge 0$.)

@@

The problem at hand is to infer the parameters of the model based on the final state of the model after it hsa evolved from some known initial state.
% If the sequence is long enough and not too much time has gone by
% this should be feasible -- there should be enough information in our observations to do this reliably.
% If the time is too large, the process may ``saturate'' --
% if there are many changes at most sites, then in at least most models,
% we will lose most information about the dynamics,
% retaining information only about the parameters that affect the stationary distribution.
% In TASEP with any boundary conditions, we lose all information as it approaches stationarity,
% while in the Ising model we have information about the temperature $\beta$ and magnetic field $\gamma$ but not the speed of changes, $\lambda$.
% In the CpG model it is not immediately clear what information is retained.
How can we extract the information?
These are Markov processes, on the state space $\calS^n$,
so the full likelihood function is given by \eqref{eqn:full_likelihood},
but doing anything with the $|\calS|^n \times |\calS|^n$ matrix $G_n$ is clearly infeasible for even moderately sized $n$.
We can, however, compute \eqref{eqn:full_likelihood} for smaller $n$,
so the first thing that one might think to do is to break the sequence up into many blocks of length $m$,
and treat these as independent.
Then, defining $p_m(t;x,y)$ to be the probability that a string $x$ of length $m$ evolves to string $y$ over time $t$,
we would obtain the approximate likelihood function
\[
  \prod_{k=0}^{n/m} p_m(x_{km+1}^{(m)},y_{km+1}^{(m)}) .
\]
However, there is no guarantee that this is even close to correct,
as it ignores dependencies between neighboring blocks.
% Intuitively, we want to increase the size of the context
% -- and the likelihood thus obtained is asymptotically correct for $m$ large,
% but computationally, we are stuck with small values of $m$.

\begin{figure}
    \begin{center}
        \includegraphics{Tmer-dependency}
    \end{center}
    \caption{
        \textbf{(A)} A $(6,2,2)$ T-mer.
        Nonmutated sites are shown in lower case letters.
        \textbf{(B)}
        The three mutations have extended the influence of the circled \nA{} 
        in the initial sequence (top)
        to the six positions colored red in the final sequence.
        Red dotted arrows show the propagation of dependency 
        between the circled \nA{}
        and the third site in the final sequence (bottom).
        The maximum window width, $w$, is equal to 1.
        \label{fig:Tmers}
    }
\end{figure}

The key insight we use is that by including neighboring sequence context in the \emph{initial} sequence only,
we can compute \emph{something} correctly,
and that something will be sufficient to compute the full likelihood.
To do this, let $p_{n,\ell,r}(x,y)$ for $r+\ell < n$ denote
the probability that an initial sequence $x$ of length $n$, after time $t$,
is found to match a smaller sequence $y$ of length $n-\ell-r$, offset by $\ell$ sites.
We refer to such patterns as ``T-mers'', as depicted in Figure \ref{fig:Tmers}.
This quantity is found by marginalizing $p_m$:
\begin{align}
    p_{n,\ell,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_n(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^n, \; y \in \calS^{n-\ell-r} ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
We expect $p_{n,\ell,m}(x,y)$ to be asymptotically correct when $\ell$ is large (and $t$ is fixed),
i.e.\ that the probability a given subsequence matching $x$ is later observed to have its center matching $y$
is close to $p_{\ell,m}(x,y)$, regardless of context wider than $2k+l$:
\begin{align} \label{eqn:window_approx}
    p_{n,\ell,r}(x,y) \approx \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}.
\end{align}

%%%%%
\subsection{Choice of overhang}

The intuition behind this fact is straightforward, and provides a way to pick the overhang lengths $\ell$ and $r$;
a precise statement of the approximation is given in Section \ref{ss:approx_pf}.
Suppose that each instantaneous change depends on positions at most $w$ sites away.
Then, the mutation process at any particular site
depends on the initial sequence at that site and within distance $w$ on either side, 
and the outcomes of any mutations that have occurred within that window.

Then, the sequence at a particular site may directly affect the process at the neighboring $w$ sites on either side.
If another change happens, say, at the end of this window,
then the indirect effects of this change may extend out to $2w-1$ sites.
In this way, the influence of the sequence at each site extends outwards, as shown in Figure \ref{fig:Tmers} --
so if we take the overhangs long enough that this dependency is unlikely to extend to the base of the Tmer,
the approximation will be good.
As we show in Section \ref{ss:approx_pf}, if the overhangs are $\ell = r = kw$,
then the error in approximation \eqref{eqn:window_approx}
is bounded by the probability that there are more than $k$ changes in a given window of length $w$.

For instance, in the CpG model the maximum dependency is $w=1$ as the process only depends on neighboring sites;
and if the sequence has evolved for time $t$ then we expect no more than a Poisson($\lambda t$) number of changes at each site.
If $\lambda t = 0.75$ (three-quarters of the sequence will have changed!),
then taking $\ell = r = 3$ allows us to compute Tmer probabilities to within an error of 0.007,
since this is the probability that a Poisson with mean 0.75 will be larger than 3.
If $\lambda t = 0.25$ we would do even better with $\ell = r = 2$.

@@

What we actually want is the \emph{full likelihood} $\like(y|x)$,
the probability that a given sequence $x$ evolves into another, $y$, over a given period of time.
This can be computed as the product of the probability of each site in $y$ given $x$ and previous sites:
defining $x_{<i}=(x_1, \ldots, x_{i-1})$, this is:
\begin{align*}
    \like(y|x) &= \prod_i \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \;\&\; X=x \} .
\end{align*}
We can compute these term by discarding sites outside the window of influence:
\begin{align*}
    \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \;\&\; X=x \} 
    &=
        \frac{
            \P\{ Y_i = y_i \;\&\; Y_{< i} = y_{<i} \;\&\; X=x \} 
        }{
            \P\{ Y_{<i} = y_{<i} \;\&\; X=x \} 
        } \\
    &\approx
        \frac{
            p_{n,\ell,\ell}(x_{i-n+\ell}^{(n)}, y_{i-n+2\ell}^{(n-2\ell)})
        }{
            p_{n,\ell,\ell+1}(x_{i-n+\ell}^{(n)}, y_{i-n+2\ell}^{(n-2\ell-1)})
        } .
\end{align*}
The approximation depends on $\ell$ being large enough, as discussed above, and $n \ge 2\ell+1$ --
this is because (a) to compute the likelihood in the numerator of the $i^\text{th}$ site, we need the flanking $\ell$ sites (as discussed above),
and (b) to compute the denominator we need to include all sites to the left affected by those $2\ell+1$ sites.

By rearranging the product, 
this gives us an expression for the full likelihood as the ratio of the likelihoods of all $(n,\ell,\ell)$ T-mers
and the likelihood of all $(n,\ell,\ell+1)$ T-mers.

@@

%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

@@

%%%%
\section*{Proof of the approximation}
\label{ss:approx_pf}

@@

In fact, the approximation improves exponentially as $\ell$ and $m$ increase --
here is a simple argument.
Suppose the maximum length of any transition pattern is $R = \max\{ |u| : (\mu,u,v) \in \calT \}$.
The transition rate at site $i$ at time $s$ is determined if we know $X_{i-R}^{(2R+1)}(s)$;
this can be determined from $X_{i-R}^{2R+1}(0)$ and the outcome of all changes happening before time $s$ at sites in $(i-R-1, \ldots, i+R)$.
These will depend in turn on the outcome of previous possible changes.
For each site $i$ and times $s_1 < s_2$, define the set of states $H(i,s_1,s_2)$
so that if one knows the possible change points $s_{ij}$,
then changes outside of $H(i,s_1,s_2)$ do not affect the distribution of $X_i(s_2)$.
Concretely, define $H(i,s_1,s_2)$ by the property that $i \in H(i,s_1,s_2)$, and
\begin{gather}
  \text{if } s_1 < s_{jk} \le s_2 \text{ for some } i-R-1 \le j \le i+R, \text{ then } H(j,s_1,s_{jk}) \subseteq H(i,s_1,s_2) .
\end{gather}
In other words, $H(i,s_1,s_2)$ is the union of the $R$-neighborhoods of $i$ and all possible changes occurring in $H(i,s_1,s_2)$ between $s_1$ and $s_2$.
For a set of states $I$, let $H(I,s_1,s_2) = \bigcup_{i\in I} H(i,s_1,s_2)$.
Formally, the state $X_i(s_2)$ is conditionally independent of $\{X_j(s_1) \st j \notin H(i,s_1,s_2)\}$.
Informally, the approximation \eqref{eqn:window_approx} will be good if $H$ does note expand beyond the flanking context.
Now,
\begin{multline}
  \P\{ X_i^{(m)}(t) = y \mid X(0) \}
  = \P\left\{ H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
   \qquad \qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\} \\
     \qquad {} + \P\left\{ H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\}  \\
   \qquad\qquad {} \times \P\left\{ X_i^{(m)}(t) = y \mid X_{i-\ell}^{(\ell+m+r)}(0), \; H(\{i,\ldots,i+m\},0,t)\nsubseteq\{i-\ell,\ldots,i+m+r\} \right\} ,
\end{multline}
and so
\begin{align} \label{eqn:prob_approx}
  \left| \P\{ X_i^{(m)}(t) = y \mid X(0) \} - p_{\ell,m}(t;x,y) \right| \le 2 \P\left\{  H(\{I,\ldots,i+m\},0,t)\subseteq\{i-\ell,\ldots,i+m+r\} \right\}.
\end{align}

@@

\subsection{s5f likelihood}

@@

\subsection{Models without using potential}

Here is a reformulation of the CpG and Ising models without using a potential.

\begin{example}[CpG mutation]
    \peter{version without selection}
  A sequence of genome can be written using A, C, G, and T;
  in the most general model of independent mutation across sites, each of the 12 possible transitions occurs at its own rate.
  Furthermore, it is a well-known observation that in many species,
  adjacent, methylated CG dinucleotides (``CpG sites'') have a much higher mutation rate to TG and CA
  than either single nucleotide change under the independent model.
  Embellishing the single-nucleotide model with this additional rate results in the model defined by
